Drought-sensitive trees survive on deep-water access in a tropical forest community

### 2020-08-30 manuscript.rmd
1. Part of changes as per co-author comments: NAte, Joe, Charlie, Stuart

- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycle, and support half of global biodiversity [@Bonan:2008; @Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]. Historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; but see @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies below-ground by vertical depth and so do rooting depths of co-occurring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018]. Consequently, trees in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically below-ground. However, such data are extremely rare [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.

\
Trade-offs shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010]. This entails identifying trade-offs among environment, traits and demography. Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011]. Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin; see Table 1 for trait symbols and their definitions). Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear. Deep rootedness may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven arid systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].

\
Huge efforts are invested in linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating that plants reach critical thresholds of hydraulic failure depending on their realized drought intensity. However, plants' realized water environments are rarely studied, especially in the tropical region. To better understand how below-ground rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, we make a thorough assessment for a tropical forest of trade-offs between the below-ground water environment, tress' above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration. Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum optimized for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductivity ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(mean(clim$Precip))` mm ($\pm$ `r round(sd(clim$Precip))` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Mean annual Penman-Monteith evapotranspiration is `r round(mean(clim$pet.PM))` mm ($\pm$ `r round(sd(clim$pet.PM))` SD; `r min(clim$Year)`-`r max(clim$Year)`). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

### Model description
\
To estimate dynamics of water availability by depth over the forest’s rooting zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions. FATES is a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].

\
Below we describe an equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013]. ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and can be described as:

\begin{equation}
\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
(\#eq:water-balance)
\end{equation}
where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step. $z$ represents the depths of exponentially distributed 15 soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).

@@

\
We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialized with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

### Model evaluation
\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)--RCT:"Matteo, please add a section"), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018). We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe ($\theta_{0-0.15 m}$). This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig. S\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 g cm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot, converted to VWC units assuming a bulk density of 0.8 g cm^-3^ ($\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019). Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m, but as depth specific calibration equation was not available, we applied the same calibration equation to these data as that for the vertical probe. However, for VWC calibration we used data from only the vertical TDR probe ($\theta_{0-0.15 m}$) to capture the long-term dynamics and the plot-wide measurements to capture the dynamics by depth. Here, in general, our interest was to capture the soil moisture dynamics by depth, so we performed model calibration in a two step process: first, we prioritized model fit with $\theta_{0.1m}$. We chose ensembles that minimized root mean square error (RMSE) between $\theta_{0.1m}$ and W~liq,`r soil.depths[8]`m~ (~300 ensembles with standaridised RMSE (range 0-1) across ensembles < 0.2). Second, we calculated an objective function for each of the 5000 ensembles by equally weighing standardized RMSE between observations and simulations across all fluxes, and chose 100 ensembles, within the 300 that best-fit $\theta_{0-0.15 m}$, that had the lowest values of the objective function. Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function can be described as

\begin{equation}
Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
(\#eq:obj-func)
\end{equation}
where $W_{liq,~\bar{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m. See Table \@ref(tab:param) for the range of best-fit values for different parameters. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

## Leaf Vulnerability Curves
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(sp.exp.param$sp)` common species (see [SI](#SIData)--RCT:"Joseph, please add a section"). We fitted four different functions to species-spefic data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011]. The differences among AICc values for these functions were not large within species. We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:

@@

## Model for Effective Rooting Depths (ERD)
\
Water is a key limiting resource for tree growth. Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:

@@

### WSG~100^$\circ$^C~ and LMA~lamina~
Joe please add, if any.

### Data gaps in WSG~100^$\circ$^C~ and LMA~lamina~

>>>>

- Among the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycle, and support half of global biodiversity [@Bonan:2008; @Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity in mitigating mortality events. Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies below-ground by vertical depth and so do rooting depths of co-occurring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018]. Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses. A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~70-90 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water with depth. However, such data are extremely rare [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.

\
Trade-offs shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010]. This entails identifying trade-offs among environment, traits and demography. Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011]. Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density; see Table 1 for trait symbols and their definitions) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear. Deep rooting may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].

\
Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. While correlations between aboveground hydraulic traits and mortality have been identified [@Anderegg:2016], numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]. However, plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region. To better understand how below-ground rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, we make a thorough assessment for a tropical forest of trade-offs between the below-ground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration. Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum optimized for increasingly reliable water availability by depth. Our hypotheses are that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductivity ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Although 28% of canopy tree species are at least partly deciduous during the dry season, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m x 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

### Model description
\
To estimate dynamics of water availability by depth over the forest’s rooting zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions. FATES is a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography approach [@Moorcroft.2001]. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].

\
Below we describe an equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013]. ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes). The equation can be described as:

\begin{equation}
\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
(\#eq:water-balance)
\end{equation}
where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step. $z$ represents the depths of 15 exponentially distributed soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).

@@

\
We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the stand structure of the 50 ha plot [@Condit.2019a] in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

### Model evaluation
\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018). We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe ($\theta_{0-0.15 m}$). This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig. S\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 g cm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot, converted to VWC units assuming a bulk density of 0.8 g cm^-3^ ($\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019). Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data as that for the vertical probe. However, for VWC calibration we used data from only the vertical TDR probe ($\theta_{0-0.15 m}$) to capture the long-term dynamics and the plot-wide measurements to capture the dynamics by depth. Here, in general, our interest was to capture the soil moisture dynamics by depth, so we performed model calibration in a two step process: first, we prioritized model fit with $\theta_{0.1m}$. We chose ensembles that minimized root mean square error (RMSE) between $\theta_{0.1m}$ and W~liq,`r soil.depths[8]`m~ (~300 ensembles with standaridised RMSE (range 0-1) across ensembles < 0.2). Second, we calculated an objective function for each of the 5000 ensembles by equally weighing standardized RMSE between observations and simulations across all fluxes, and chose 100 ensembles, within the 300 that best-fit $\theta_{0-0.15 m}$, that had the lowest values of the objective function. Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function can be described as

\begin{equation}
Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
(\#eq:obj-func)
\end{equation}
where $W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m. See Table \@ref(tab:param) for the range of best-fit values for different parameters. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

## Leaf Vulnerability Curves
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(sp.exp.param$sp)` common species (see [SI](#SIData)). We fitted four different functions to species-specific data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011]. The differences among AICc values for these functions were not large within species (Zailaa et al. to be submitted). We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:

@@

## Model for Effective Rooting Depths (ERD)
\
Water is a key limiting resource for tree growth. Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:

@@

### Data gaps in WSG and LMA
